---
title: "Week 2 - Homework"
author: "STAT 420, Summer 2018, Bryna Zhao"
date: ''
output:
  html_document: 
    toc: yes
  pdf_document: default
urlcolor: cyan
---

***

## Exercise 1 (Using `lm`)

For this exercise we will use the `cats` dataset from the `MASS` package. You should use `?cats` to learn about the background of this dataset.

**(a)** Suppose we would like to understand the size of a cat's heart based on the body weight of a cat. Fit a simple linear model in `R` that accomplishes this task. Store the results in a variable called `cat_model`. Output the result of calling `summary()` on `cat_model`.

```{r}
library(MASS)
#View(cats)
cat_model = lm(Hwt ~ Bwt, data = cats)
summary(cat_model)
```

**(b)** Output only the estimated regression coefficients. Interpret $\hat{\beta_0}$ and $\beta_1$ in the *context of the problem*. Be aware that only one of those is an estimate.

**$\hat{\beta_0}$ or the intercept tells us the estimated mean heart weight for a cat with a body weight of 0kg is -0.3566624 g. It's just an estimate and not likely to happen. And $\beta_1$ or the slope indicates that for every additional kg in cat's body weight we could expect heart weight to increase by an average of 4.0340627 g.**

```{r}
coef(cat_model)
```


**(c)** Use your model to predict the heart weight of a cat that weights **2.7** kg. Do you feel confident in this prediction? Briefly explain.

**A cat weighting 2.7 kg is predicted to have a heart weight of 10.53531 g.Yes, I feel confident. Since 2.7 is within the data range (2.0 ~ 3.9), it's considered interpoloation.**

```{r}
predict(cat_model, newdata = data.frame(Bwt = 2.7))
range(cats$Bwt)
min(cats$Bwt) < 2.7 & 2.7 <max(cats$Bwt)
```


**(d)** Use your model to predict the heart weight of a cat that weights **4.4** kg. Do you feel confident in this prediction? Briefly explain.

**A cat weighting 4.4 kg is predicted to have a heart weight of 17.39321 g. No, I feel less confident in this prediction. Since 4.4 is not an observed value of Bwt and is outside the data range (2.0 ~ 3.9), it's considered extrapoloation.**

```{r}
predict(cat_model, newdata = data.frame(Bwt = 4.4))
min(cats$Bwt) < 4.4 & 4.4 <max(cats$Bwt)
```


**(e)** Create a scatterplot of the data and add the fitted regression line. Make sure your plot is well labeled and is somewhat visually appealing.
```{r}
plot(Hwt ~ Bwt, data = cats,
     xlab = "Cat Body Weight in KG",
     ylab = "Cat Heart Weight in G",
     main = "Cat Body Weights vs. Heart Weights",
     pch = 20,
     cex = 2,
     col = "aquamarine3")
abline(cat_model, lwd = 3, col = "brown")

```

**(f)** Report the value of $R^2$ for the model. Do so directly. Do not simply copy and paste the value from the full output in the console after running `summary()` in part **(a)**.

**$R^2$ of this model is 0.6466209**
```{r}
summary(cat_model)$r.squared
```
```{r}
x = cats$Bwt
y = cats$Hwt
Sxy = sum((x - mean(x)) * (y - mean(y)))
Sxx = sum((x - mean(x)) ^ 2)
beta_1_hat = Sxy / Sxx
beta_0_hat = mean(y) - beta_1_hat * mean(x)
y_hat = beta_0_hat + beta_1_hat * x
SST = sum((y - mean(y)) ^ 2)
SSE = sum((y - y_hat) ^ 2)
1 - SSE / SST
```


***

## Exercise 2 (Writing Functions)

This exercise is a continuation of Exercise 1.

**(a)** Write a function called `get_sd_est` that calculates an estimate of $\sigma$ in one of two ways depending on input to the function. The function should take three arguments as input:

- `fitted_vals` - A vector of fitted values from a model
- `actual_vals` - A vector of the true values of the response
- `mle` - A logical (`TRUE` / `FALSE`) variable which defaults to `FALSE`

The function should return a single value:

- $s_e$ if `mle` is set to `FALSE`.
- $\hat{\sigma}$ if `mle` is set to `TRUE`.

```{r}
get_sd_est = function(fitted_vals, actual_vals, mle = FALSE){
  if (! mle){
    s2_e = sum((actual_vals - fitted_vals) ^ 2 ) / (length(actual_vals) - 2)
    s_e = sqrt(s2_e)
    return(s_e)
  } 
  else{
    sigma2_hat = sum((actual_vals - fitted_vals) ^ 2 ) / length(actual_vals)
    sigma_hat = sqrt(sigma2_hat)
    return(sigma_hat)
  }

}



  
```


**(b)** Run the function `get_sd_est` on the residuals from the model in Exercise 1, with `mle` set to `FALSE`. Explain the resulting estimate in the context of the model.
**When mle is set to False, we're using the least squares approach. $s_e$ or the residual standard error, is the standard deviation of the residuals and it tells us our estimates of mean cat heart weight vary from the actual average value of the response variable by 1.452373 g. **

```{r}
get_sd_est(cat_model$fitted.values, cats$Hwt)
```


**(c)** Run the function `get_sd_est` on the residuals from the model in Exercise 1, with `mle` set to `TRUE`. Explain the resulting estimate in the context of the model. Note that we are trying to estimate the same parameter as in part **(b)**.
**When mle is set to TRUE, we're using the maximum likelihood estimates and have a different estimate of $\hat{\sigma}$ of 1.442252. According to this estimate, our estimated mean of cat heart weight vary from the actual average value of the response variable by 1.442252 g. This estimate is biased. **

```{r}
get_sd_est(cat_model$fitted.values, cats$Hwt, mle = TRUE)
```


**(d)** To check your work, output `summary(cat_model)$sigma`. It should match at least one of **(b)** or **(c)**.
**The sigma matches the result of b.**

```{r}
summary(cat_model)$sigma
all.equal(get_sd_est(cat_model$fitted.values, cats$Hwt), summary(cat_model)$sigma)
```

***

## Exercise 3 (Simulating SLR)

Consider the model

\[
Y_i = 5 + -3 x_i + \epsilon_i
\]

with 

\[
\epsilon_i \sim N(\mu = 0, \sigma^2 = 10.24)
\]

where $\beta_0 = 5$ and $\beta_1 = -3$.

This exercise relies heavily on generating random observations. To make this reproducible we will set a seed for the randomization. Alter the following code to make `birthday` store your birthday in the format: `yyyymmdd`. For example, [William Gosset](https://en.wikipedia.org/wiki/William_Sealy_Gosset), better known as *Student*, was born on June 13, 1876, so he would use:

```{r}
birthday = 19870725
set.seed(birthday)
```

**(a)** Use `R` to simulate `n = 25` observations from the above model. For the remainder of this exercise, use the following "known" values of $x$.

```{r}
beta_0 = 5
beta_1 = -3
sigma = sqrt(10.24)
x = runif(n = 25, 0, 10)
sim_slr = function(x, beta_0, beta_1, sigma) {
  n = length(x)
  epsilon = rnorm(n, mean = 0, sd = sigma)
  y = beta_0 + beta_1 * x + epsilon
  data.frame(predictor = x, response = y)
}

sim_data = sim_slr(x, beta_0 = 5, beta_1 =-3, sigma = sqrt(10.24))
sim_data$predictor
sim_data$response

```

You may use [the `sim_slr ` function provided in the text](http://daviddalpiaz.github.io/appliedstats/simple-linear-regression.html#simulating-slr). Store the data frame this function returns in a variable of your choice. Note that this function calls $y$ `response` and $x$ `predictor`.

**(b)** Fit a model to your simulated data. Report the estimated coefficients. Are they close to what you would expect? Briefly explain.

**Yes, true value of coefficients are 5 and -3, the estimates our model are 3.734827 and  -2.705183. Our estimates are not too far off the true values. **

```{r}
sim_model = lm(sim_data$response ~ sim_data$predictor)
coef(sim_model)
```


**(c)** Plot the data you simulated in part **(a)**. Add the regression line from part **(b)** as well as the line for the true model. Hint: Keep all plotting commands in the same chunk.

```{r}

plot(response ~ predictor, data = sim_data,
     xlab = "X Values",
     ylab = "Estimated Y Values",
     main = "Simulated data",
     pch = 20,
     cex = 2,
     col = "grey",
     ylim = c(-20,10))
abline(sim_model, lwd = 3, col = "brown")
abline(beta_0, beta_1, lwd = 3, lty = 2, col = "dodgerblue")
legend("topright", c("Estimate", "Truth"), lty = c(1, 2), lwd = 2,
       col = c("brown", "dodgerblue"))
```


**(d)** Use `R` to repeat the process of simulating `n = 25` observations from the above model $1500$ times. Each time fit a SLR model to the data and store the value of $\hat{\beta_1}$ in a variable called `beta_hat_1`. Some hints:

- Consider a `for` loop.
- Create `beta_hat_1` before writing the `for` loop. Make it a vector of length $1500$ where each element is `0`.
- Inside the body of the `for` loop, simulate new $y$ data each time. Use a variable to temporarily store this data together with the known $x$ data as a data frame.
- After simulating the data, use `lm()` to fit a regression. Use a variable to temporarily store this output.
- Use the `coef()` function and `[]` to extract the correct estimated coefficient.
- Use `beta_hat_1[i]` to store in elements of `beta_hat_1`.
- See the notes on [Distribution of a Sample Mean](http://daviddalpiaz.github.io/appliedstats/introduction-to-r.html#distribution-of-a-sample-mean) for some inspiration.

You can do this differently if you like. Use of these hints is not required.

```{r}
x = runif(n = 25, 0, 10)

run_sim = function(x_val, beta0_val, beta1_val, sigma_val, size){
  beta_hat_vec = rep(0, times = size)
  

  for (i in 1:size){
  y = sim_slr(x = x_val, beta_0 = beta0_val, beta_1 = beta1_val, sigma = sigma_val)
  sim_model = lm(y$response ~ y$predictor)
  beta_hat_vec[i] = coef(sim_model)[[2]]
  }
  
  return(beta_hat_vec)
}
beta_hat_1 = run_sim(x_val = x, beta0_val = 5, beta1_val = -3, sigma_val = sqrt(10.24), 1500)
# beta_hat_1


```


**(e)** Report the mean and standard deviation of `beta_hat_1`. Do either of these look familiar?
**The mean beta_hat_1 is very close to -3, which is the true $\beta_1$.**

```{r}
mean(beta_hat_1)
sd(beta_hat_1)
```


**(f)** Plot a histogram of `beta_hat_1`. Comment on the shape of this histogram.
**The histogram has a shape of normal distribution.**
```{r}
hist(beta_hat_1,
     xlab = "Mean of Beta_1 of Simulated Data",
     main = "Mean of Beta_1 in 1500 Times of Simulation",
     breaks = 100,
     xlim = c(-4,-2),
     col = "dodgerblue")
```


***

## Exercise 4 (Be a Skeptic)

Consider the model

\[
Y_i = 3 + 0 \cdot x_i + \epsilon_i
\]

with

\[
\epsilon_i \sim N(\mu = 0, \sigma^2 = 4)
\]

where $\beta_0 = 3$ and $\beta_1 = 0$.

Before answering the following parts, set a seed value equal to **your** birthday, as was done in the previous exercise.

```{r}
birthday = 19870725
set.seed(birthday)
```

**(a)** Use `R` to repeat the process of simulating `n = 75` observations from the above model $2500$ times. For the remainder of this exercise, use the following "known" values of $x$.

```{r}
x = runif(n = 75, 0, 10)
beta_hat_1 = rep(0, times = 2500)
beta_hat_1 = run_sim(x_val = x, beta0_val = 3, beta1_val = 0, sigma_val = sqrt(4), 2500)

```

Each time fit a SLR model to the data and store the value of $\hat{\beta_1}$ in a variable called `beta_hat_1`. You may use [the `sim_slr ` function provided in the text](http://daviddalpiaz.github.io/appliedstats/simple-linear-regression.html#simulating-slr). Hint: Yes $\beta_1 = 0$.

**(b)** Plot a histogram of `beta_hat_1`. Comment on the shape of this histogram.
**The histogram has a shape of normal distribution.**
```{r}
hist(beta_hat_1,
     xlab = "Mean of Beta_1 of Simulated Data",
     main = "Mean of Beta_1 in 2500 Times of Simulation",
     breaks = 100,
     xlim = c(-0.3, 0.3),
     col = "darkblue")
```


**(c)** Import the data in [`skeptic.csv`](skeptic.csv) and fit a SLR model. The variable names in `skeptic.csv` follow the same convention as those returned by `sim_slr()`. Extract the fitted coefficient for $\beta_1$.
```{r}
library(readr)
skeptic <- read_csv("skeptic.csv")
# View(skeptic)
slr_model = lm(response ~ predictor, data = skeptic)
beta_1_hat = coef(slr_model)[[2]]
beta_1_hat

```


**(d)** Re-plot the histogram from **(b)**. Now add a vertical red line at the value of $\hat{\beta_1}$ in part **(c)**. To do so, you'll need to use `abline(v = c, col = "red")` where `c` is your value.

```{r}
hist(beta_hat_1,
     xlab = "Mean of Beta_1 of Simulated Data",
     main = "Mean of Beta_1 in 2500 Times of Simulation",
     breaks = 100,
     xlim = c(-0.3, 0.3),
     col = "darkblue")
abline(v = beta_1_hat, col = "red")
```


**(e)** Your value of $\hat{\beta_1}$ in **(c)** should be negative. What proportion of the `beta_hat_1` values is smaller than your $\hat{\beta_1}$? Return this proportion, as well as this proportion multiplied by `2`.

**0.08% of `beta_hat_1` is smaller than the `beta_hat_1`.**
```{r}
sum(beta_hat_1 < beta_1_hat) / length(beta_hat_1)
sum(beta_hat_1 < beta_1_hat) / length(beta_hat_1) * 2
```


**(f)** Based on your histogram and part **(e)**, do you think the [`skeptic.csv`](skeptic.csv) data could have been generated by the model given above? Briefly explain.
**No, the chance is very small, since the beta from the skeptic data is too far off the mean of the normal distribution  and there is only 0.08% possibility that the skeptic data could have been generated by the model given above.**


***

## Exercise 5 (Comparing Models)

For this exercise we will use the `Ozone` dataset from the `mlbench` package. You should use `?Ozone` to learn about the background of this dataset. You may need to install the `mlbench` package. If you do so, do not include code to install the package in your `R` Markdown document.

For simplicity, we will perform some data cleaning before proceeding.
```{r eval=FALSE}
install.packages("mlbench")
```


```{r}

data(Ozone, package = "mlbench")
Ozone = Ozone[, c(4, 6, 7, 8)]
colnames(Ozone) = c("ozone", "wind", "humidity", "temp")
Ozone = Ozone[complete.cases(Ozone), ]
```

We have:

- Loaded the data from the package
- Subset the data to relevant variables
    - This is not really necessary (or perhaps a good idea) but it makes the next step easier
- Given variables useful names
- Removed any observation with missing values
    - This should be given much more thought in practice

For this exercise we will define the "Root Mean Square Error" of a model as

\[
\text{RMSE} = \sqrt{\frac{1}{n} \sum_{i = 1}^{n}(y_i - \hat{y}_i)^2}.
\]

**(a)** Fit three SLR models, each with "ozone" as the response. For the predictor, use "wind speed," "humidity percentage," and "temperature" respectively. For each, calculate $\text{RMSE}$ and $R^2$. Arrange the results in a markdown table, with a row for each model. Suggestion: Create a data frame that stores the results, then investigate the `kable()` function from the `knitr` package.

```{r eval=FALSE}
# if you haven't install kableExtra
install.packages("kableExtra")  

```


```{r}
# ozone vs wind
rmse_fc = function(y, y_hat){
  rmse = sqrt(sum((y - y_hat) ^ 2 ) / length(y))
  return(rmse)
}
wind_model = lm(ozone ~ wind, data = Ozone)
wind_rmse = rmse_fc(Ozone$ozone, wind_model$fitted.values)
wind_R2 = summary(wind_model)$r.squared

# ozone vs humidity
humid_model = lm(ozone ~ humidity, data = Ozone)
humid_rmse = rmse_fc(Ozone$ozone, humid_model$fitted.values)
humid_R2 = summary(humid_model)$r.squared

# ozone vs temp
temp_model = lm(ozone ~ temp, data = Ozone)
temp_rmse = rmse_fc(Ozone$ozone, temp_model$fitted.values)
temp_R2 = summary(temp_model)$r.squared

# Form the table  
wind = c(wind_rmse, wind_R2)
humid = c(humid_rmse, humid_R2)
temp = c(temp_rmse, temp_R2)

compare_table = rbind(wind, humid, temp)
library(knitr)
library(kableExtra)
compareThree = kable(compare_table, format = "html", col.names = c("Root Mean Square Error", "R Squared"),caption = "Compare Three Models") %>% kable_styling("striped")
compareThree


```


**(b)** Based on the results, which of the three predictors used is most helpful for predicting wins? Briefly explain.

**The temp model definitely wins. Because it's achieved a relatively high r-squared and a relatively low root mean square error. The root mean square error is measuring how far the average response deviate from the true regression line, so we want it low while r-squared is indicating how well a model is fitting the actual data, so we want it high. **

***

## Exercise 00 (SLR without Intercept)

**This exercise will _not_ be graded and is simply provided for your information. No credit will be given for the completion of this exercise. Give it a try now, and be sure to read the solutions later.**

Sometimes it can be reasonable to assume that $\beta_0$ should be 0. That is, the line should pass through the point $(0, 0)$. For example, if a car is traveling 0 miles per hour, its stopping distance should be 0! (Unlike what we saw in the book.)

We can simply define a model without an intercept,

\[
Y_i = \beta x_i + \epsilon_i.
\]

**(a)** [In the **Least Squares Approach** section of the text](http://daviddalpiaz.github.io/appliedstats/simple-linear-regression.html#least-squares-approach) you saw the calculus behind the derivation of the regression estimates, and then we performed the calculation for the `cars` dataset using `R`. Here you need to do, but not show, the derivation for the slope only model. You should then use that derivation of $\hat{\beta}$ to write a function that performs the calculation for the estimate you derived. 

In summary, use the method of least squares to derive an estimate for $\beta$ using data points $(x_i, y_i)$ for $i = 1, 2, \ldots n$. Simply put, find the value of $\beta$ to minimize the function

\[
f(\beta)=\sum_{i=1}^{n}(y_{i}-\beta x_{i})^{2}.
\]

Then, write a function `get_beta_no_int` that takes input:

- `x` - A predictor variable
- `y` - A response variable

The function should then output the $\hat{\beta}$ you derived for a given set of data.
```{r}
get_beta_no_int = function(x, y){
  beta = sum(x * y) / sum(x ^ 2)
  return(beta)
}
```


**(b)** Write your derivation in your `.Rmd` file using TeX. Or write your derivation by hand, scan or photograph your work, and insert it into the `.Rmd` as an image. See the [RMarkdown documentation](http://rmarkdown.rstudio.com/) for working with images.



**(c)** Test your function on the `cats` data using body weight as `x` and heart weight as `y`. What is the estimate for $\beta$ for this data?
```{r}
get_beta_no_int(cats$Bwt, cats$Hwt)
```


**(d)** Check your work in `R`. The following syntax can be used to fit a model without an intercept:

```{r, eval = FALSE}
lm(response ~ 0 + predictor, data = dataset)
```

Use this to fit a model to the `cat` data without an intercept. Output the coefficient of the fitted model. It should match your answer to **(c)**.

```{r}
coef(lm(Hwt ~ 0 + Bwt, data = cats))
```


